Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-20 manuscript/response_to_reviews.Rmd

Statistical analysis: the analysis would be more precise if the authors used Generalized linear mixed models to avoid data transformation and take into account (with random effects, as they do for the linear mixed model, LMM, presented now) the influence of species in the analysis. Then I missed a further discussion on the role of individual species (e.g. see my comment below on Rt>1).

**A GLMM would only avoid transformation of the dependent variable (Rt), but that was (and is) not transformed. We tried the analysis with generalized linear mixed models with a Gaussian distribution, and confirmed that this gave very similar results, so we are sticking with LMM.**

I suggest to show some relationship, e.g. between Rt and traits in figures. The study only presents two figures and there is plenty of room to present results more clearly with extra figures.

**We created two new figures: one showing how $Rt$ varied across species in each of the three droughts (Fig. 3) and one visualizing the best models (Fig. 4).**

### 2020-07-20 manuscript/climate_sensitivity_draft.Rmd
Bourg comments
#110

Many studies have shown that within species, large trees tend to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @hacket-pain_consistent_2016). It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. First, tree height itself may be a primary driver. Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018]. Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al. 2010, Olson et al. 2014,2018*). Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 3). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data are available through the ForestGEO data portal (www.forestgeo.si.edu).

@@

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019]. Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **2**). There was no evidence that greater availability of, or access to soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits, consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought ([*Stovall et al. 2020*](https://doi.org/10.1038/s41467-020-17213-5)). Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$). This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The directions of these responses were consistent across droughts (Table S6), supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

@@

### Author Contribution 
KAT, IM, and AJT designed the research. Tree-ring chronologies were developed by RH under guidance of AJT and NP. Trait data was collected by IM, JZ under guidance of NK and LS. Other plot data were collected by IM, AS, EGA, and NB under guidance of EGA and WM. Data analyses were performed by IM under guidance of KAT and VH. KAT and IM interpreted the results. IM and KAT wrote the first draft of manuscript, and all authors contributed to revisions.

>>>>

Many studies have shown that within species, large trees tend to be more affected by drought. Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @hacket-pain_consistent_2016). It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. First, tree height itself may be a primary driver. Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018]. Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018]. Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al. 2010, Olson et al. 2014,2018*). Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 3). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From these census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data are available through the ForestGEO data portal (www.forestgeo.si.edu).

@@

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019]. Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **2**). There was no evidence that greater availability of, or access to soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits, consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought ([*Stovall et al. 2020*](https://doi.org/10.1038/s41467-020-17213-5)). Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$). This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The directions of these responses were consistent across droughts (Table S6), supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought -- and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

@@

### Author Contribution 
KAT, IM, and AJT designed the research. Tree-ring chronologies were developed by RH under guidance of AJT and NP. Trait data were collected by IM, JZ under guidance of NK and LS. Other plot data were collected by IM, AS, EGA, and NB under guidance of EGA and WM. Data analyses were performed by IM under guidance of KAT and VH. KAT and IM interpreted the results. IM and KAT wrote the first draft of manuscript, and all authors contributed to revisions.
